\section{Case Study: Proving Lemmas for the Kepler Conjecture}\label{kepler}

We build the proof-producing capacity into our open-source tool {\sf
  dReal}\footnote{ \url{http://dreal.cs.cmu.edu} }. All the
experiments below are run on a machine of with a 32-core 2.0GHz Intel
Xeon E5-2600 Processor and 64GB of RAM.

\input{good_proofs.tex}

A main motivation for us to build the proof checker is to contribute
to the Flyspeck project~\cite{DBLP:conf/dagstuhl/Hales05}, for the
fully formalized proof of the Kepler conjecture. As lemmas for the
proof, hundreds of nonlinear real inequalities need to be verified.
Although the formulas usually contain only around ten variables, they
contain a huge number of nonlinear arithmetic operations and
trigonometric functions, and are mathematically challenging. The
following is typical:

\begin{align*}
\forall \vec x&\in [2, 2.51]^6.\ \Big( -\frac{\pi-4\arctan\frac{\sqrt
2}{5}}{12\sqrt2}\sqrt{\Delta(\vec x)}\\
&+\frac{2}{3}\sum_{i=0}^3\arctan\frac{\sqrt{\Delta(\vec x)}}{a_i(\vec x)}\leq
-\frac{\pi}{3}+4\arctan\frac{\sqrt 2}{5}\Big)
\end{align*}

where $a_i(\vec x)$ are quadratic and $\Delta(\vec x)$ is the
determinant of a nonlinear matrix.

In the original proof, Hales implemented procedures that combine
linear programming and interval arithmetic to establish all these
formulas, but the algorithms are hard to be formally verified. In
fact, the formal verification of these formulas is the last main piece
of work needed to complete the full project. The state of the art is
explained in a very recent thesis~\cite{keplerthesis}, reporting the
proofs of about 10 inequalities so far, using formal Taylor series and
interval arithmetic.

Without any particular optimization on ICP, we have already observed
promising results. Out of 916 nonlinear formulas in the Flyspeck
project repository, we solved 111 of them returning {\sf unsat} with a
timeout of 180 seconds and precision $\delta=10^{-3}$. Out of these
formulas, we applied the proof checking algorithm, and formally proved
72 of them directly.The proof traces of these formulas can be very
large; for instance, we proved one with more than 2M lines in the
proof (54MB file). In Table~\ref{tbl:exp}, we list some of the
representative benchmarks to show scalability. A full table for all
the results is on the tool page. It also contains results for other
standard benchmarks, such as from~\cite{DBLP:conf/cade/JovanovicM12}.

As the proof-generation capacity is for debugging SMT solvers, the
proofs have also been valuable for us to observe bugs during the
process. For instance, we have observed unstable behavior in the
trigonometric function evaluation in {\sf realpaver} (the cosine
function around $4\pi/3$). On the other hand, our proof checker only
involves simple operations written in {\sf OCaml}, and performs the
checking completely independently from the solver. Thus, for the {\sf
  unsat} answers with proofs, we are confident about the correctness
of the answer, which does not involve bugs from {\sf realpaver}.
Ultimately, we aim for having a verified proof checker in systems such
as HOL.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
