\section{Bounded $\delta$-reachability}\label{sec:delta-reachability}
Let $H = \langle X, Q, \flow, \jump, \inv, \init\rangle$ be a hybrid
system as standardly defined. We use first-order formulae over the
real numbers to represent $H$, by writing $$H = \langle X, Q,
\varphi_{\flow}, \varphi_{\jump}, \varphi_{\inv},
\varphi_{\init}\rangle$$ where $\varphi_{\flow}, \varphi_{\jump},
\varphi_{\inv}$ and $\varphi_{\init}$ are logic formulae that define
the corresponding predicates in the standard definition. Now, let
$\delta\in \mathbb{Q}^+$ be a chosen error bound, we define the
$\delta$-perturbation of $H$ to be $$H^{\delta} = \langle X, Q,
\varphi_{\flow}^{\delta}, \varphi_{\jump}^{\delta},
\varphi_{\inv}^{\delta}, \varphi_{\init}^{\delta}\rangle.$$ Here,
$\varphi^{\delta}$ is a syntactic variant of $\varphi$ which relaxes
the numerical terms in $\varphi$ up to an error bound $\delta$. The
notion is formally defined in our recent
work~\cite{DBLP:conf/lics/GaoAC12,DBLP:conf/cade/GaoAC12,DBLP:journals/corr/GaoKCC14}.
We now define the bounded $\delta$-reachability problem that \dReach{}
solves.

%\paragraph{Bounded $\delta$-Reachability.}
Let $n\in \mathbb{N}$ be a bound and $T\in \mathbb{R}^+$ be an upper
bound of time duration. We write $\unsafe$ to denote a subset of the
state space of $H$ defined by a first-order formula. The bounded
$\delta$-reachability problem asks for one of the following answers
\begin{itemize}
 \item $H$ cannot reach $\unsafe$ in $n$ steps within time $T$.
 \item $H^{\delta}$ can reach $\unsafe^{\delta}$ in $n$ steps within time $T$.
\end{itemize}
Note that these answers are not weaker than the precise ones. When
{\sf safe} is the answer, we know for certain that $H$ does not reach
the unsafe region; when {\sf $\delta$-unsafe} is the answer, there
exists some $\delta$-bounded perturbation in the system that {\em
  would} render it unsafe. Note that the error-bound $\delta$ can be
chosen to be arbitrarily small, so that the {\sf$\delta$-unsafe}
answer discovers robustness problem in the system, which should be
regarded as unsafe indeed.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
