%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}
\input{macros}

\begin{document}

\permission{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from Permissions@acm.org. }
\conferenceinfo{HSCC'15}{April 14 - 16, 2015, Seattle, WA, USA}
\copyrightetc{Copyright 2015 ACM \the\acmcopyr}
\crdata{978-1-4503-3433-4/15/04\$15.00\\
http://dx.doi.org/10.1145/2728606.2728634}

\title{Towards Personalized Prostate Cancer Therapy Using Delta-Reachability Analysis}%\titlenote{This work has been partially supported by award N00014-13-1-0090 of the US Office of Naval Research and award CNS0926181 of the National Science foundation (NSF).}}


\numberofauthors{6} 
\author{
% 1st. author
\alignauthor
Bing Liu\\
       \affaddr{School of Medicine}\\
       \affaddr{University of Pittsburgh}\\
       %\affaddr{Pittsburgh, PA, USA}\\
       \email{liubing@pitt.edu}
% 2nd. author
\alignauthor
Soonho Kong\\
       \affaddr{Computer Science Dept.}\\
       \affaddr{Carnegie Mellon University}\\
      % \affaddr{Pittsburgh, PA, USA}\\
       \email{soonhok@cs.cmu.edu}
% 3rd. author
\alignauthor 
Sicun Gao\\
       \affaddr{CSAIL}\\
       \affaddr{MIT}\\
       %\affaddr{Cambridge, MA, USA}\\
       \email{sicung@csail.mit.edu}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor 
Paolo Zuliani\\
       \affaddr{School of Computer Science}\\
       \affaddr{Newcastle University}\\
      % \affaddr{Newcastle, UK}\\
       \email{paolo.zuliani@ncl.ac.uk}
% 5th. author
\alignauthor 
Edmund M. Clarke\\
       \affaddr{Computer Science Dept.}\\
       \affaddr{Carnegie Mellon University}\\
       %\affaddr{Pittsburgh, PA, USA}\\
       \email{emc@cs.cmu.edu}
}

\date{20 February, 2015}


\maketitle

\begin{abstract}
Recent clinical studies suggest that the efficacy of hormone therapy for prostate cancer depends on the characteristics of individual patients. In this paper, we develop a computational framework for identifying patient-specific androgen ablation therapy schedules for postponing the potential cancer relapse. We model the population dynamics of heterogeneous prostate cancer cells in response to androgen suppression as a nonlinear hybrid automaton. We estimate personalized kinetic parameters to characterize patients and employ $\delta$-reachability analysis to predict patient-specific therapeutic strategies. The results show that our methods are promising and may lead to a prognostic tool for prostate cancer therapy.
\end{abstract}


\category{D.2.4}{Software Engineering}{Software/Program Verification}[Model checking]
\category{J.3}{Life and Medical Sciences}{Biology and genetics}

\terms{Theory, Verification}

\keywords{hybrid systems, delta-reachability, systems biology, prostate cancer, personalized therapy}


\input{intro}
\input{model}
\input{method}
\input{results}
\input{conclusion}




%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This work has been partially supported by award N00014-13-1-0090 of the US Office of Naval Research and award CNS0926181 of the National Science foundation (NSF).


\bibliographystyle{abbrv}
\bibliography{sigproc}

%\input{appendix}

\end{document}
